\documentclass[10pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[top=1in, bottom=1.25in, left=1.25in, right=1.25in]{geometry}
\usepackage{amsmath}
\usepackage[round]{natbib}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\author{Benjamin Peter, Arev}

\newcommand{\BE}[1]{\mathbb{E}\left[#1\right]}
\newcommand{\BFZ}{\mathbf{Z}}
\newcommand{\BFG}{\mathbf{G}}
\newcommand{\BFO}{\mathbf{O}}
\newcommand{\BFC}{\mathbf{C}}
\newcommand{\BFX}{\mathbf{X}}

\title{Supplement 1.3: Model Details}
\begin{document}
	\maketitle
\section*{Model Overview}
In this section, we give the full mathematical details of the admixslug model and its implementation.
	
\subsection*{Notation overview}
	Throughout, the notation is as follows: 
	\begin{itemize}
		\item $R$ is the number of read groups
		\item $L$ is the number of biallelic SNP loci
		\item $K$ is the number of SFS-bins, which are entries in the reference SFS. 
		\item $n_{rl}$ the number of reads of read group $r$ at SNP $l$.
		\item $O_{rli}$ is the $i$-th read from read group $r$ at SNP $l$
		\item $G_{l}$ is the genotype at SNP $l$, codified by the number of derived alleles
		\item $\psi_l$ is the derived allele frequency in a contaminant panel at locus $l$
		\item $Z_{l}$ is the set of all SFS entries, and the SFS-entry corresponding to SNP $l$
		\item $X_{rli}$ is an indicator that is one if the true allele corresponding to $O_{rli}$ is derived, zero if it is ancestral
		\item $C_{rli}$ an indicator that is 1 if the read corresponding to $X_{rli}$ is of contaminant origin, and 0 if it is endogenous
		\item $F_k$ a parameter estimating coalescence since gene flow for SNP in SFS-entry $k$.
		\item $\tau_k$ the proportion of derived alleles in SFS-entry $k$.		
		\item $c_r$ proportion of contaminant reads in read group $r$
		\item $e,b$  error rate, and reference bias
		\item $W_l$ 
		\item $\mathbf{\theta} = (c_r, e, b, \tau_k, F_k)$, the set of all parameters to be estimated
	\end{itemize}
In all cases, we use a bold capital letter to denote the set of all parameters, i.e. $\BFG = (G_l)$ is the set of all genotypes.

\section*{Model details}
\subsection*{Error model}
The error model relates the (unobserved) true state of the sequenced molecule $X_{lri}$ to the observed sequencing data $O_{lri}$, which denotes the $i$-th read of read group $r$ at locus $l$. We consider sequencing error,and reference bias.

$X$ is based on ancestral/derived alleles, whereas $O$ is codified by reference/alternative alleles. We further have a variable $W_l$ that is 1 if the reference allele is flipped, i.e.

\begin{align*}
W_l = \begin{cases}
0 & \text{ if REF = Ancestral allele}\\
1&  \text{ if REF = Derived allele}
\end{cases}
\end{align*}

The eight possible cases are then:

\begin{table}[!ht]
\begin{center}

\begin{tabular}{r r r l | r}
Observed base & base on seq & flipped & alleles & probability\\
$O_{rli}$ & $X_{rli}$ & $W_l$ &\\
\hline
0 (ref) & 0 (anc)& 0 &ref = anc, alt = der & 1 - e\\
1 (alt) & 0 (anc)& 0 &ref = anc, alt = der& e\\
0 (ref) & 1 (der)& 0 &ref = anc, alt = der& b\\
1 (alt) & 1 (der)& 0 &ref = anc, alt = der& 1 - b\\
0 (ref) & 0 (anc)& 1 &ref = der, alt = anc& b \\
1 (alt) & 0 (anc)& 1 &ref = der, alt = anc& 1 - b\\
0 (ref) & 1 (der)& 1 &ref = der, alt = anc& 1 - e\\
1 (alt) & 1 (der)& 1 &ref = der, alt = anc& e \\
\end{tabular}
\end{center}
\end{table}


We can think of $e$ as the sequencing error, and $b$ as the reference bias + sequencing error.

This is implemented in \texttt{bwd\_p\_o\_given\_x}, which calculates a matrix of size $R \times 2$ containing the entries
$P(O_i | X_i = j)$, flattened over $r$ and $l$

\subsection*{Sequence model}

There are two possible origins for each sequence $X_{lri}$, it is either a contaminant, or endogenous. 
Let $C_{lri} =1 $ mean that $X_{lri}$ is contaminant, and $C_{lri} =0 $ mean it is not. Furthermore, let $\psi_l$ be the \textbf{derived} allele frequency in a reference contamination panel, which we assume to be known. Let $G_l$ be the genotype of the target individual at SNP $l$, counted as the number of derived alleles (which is either 0, 1 or 2 in diploid individuals).


\begin{align}
P(X_{lri} = 0 | C_{lri} = 1, \psi_l) &= 1 - \psi_l\nonumber\\ 
P(X_{lri} = 1 | C_{lri} = 1, \psi_l) &= \psi_l\nonumber\\ 
P(X_{lri} = 0 | C_{lri} = 0, G_l) &= 1 - \frac{G_l}{2}\nonumber\\ 
P(X_{lri} = 1 | C_{lri} = 0, G_l) &= \frac{G_l}{2}
\end{align}

In haploid regions (or more generally, in regions with different ploidy) we divide $G_l$ by the ploidy instead.

%\paragraph{Faunal contamination}
%For faunal contamination, we might expect always the ancestral allele. Let $\phi_l=0$ mean the reference allele at locus $l$ is ancestral, in which case faunal contamination would have the reference allele. Likewise, if the derived allele is ancestral, then  $\phi_l=1$. 

%If we add faunal contamination, we extend $C_{lri}$ to an additional state, and add
%\begin{align}
%P(X_{lri} = 0 | C_{lri} = 2, \phi_l) &= 1-\phi_l\nonumber\\ 
%P(X_{lri} = 1 | C_{lri} = 2, \phi_l) &= \phi_l
%\end{align}

\subsection*{Contamination model}
For read-group $r$, the probability that a read from that read group is a contaminant is 
\begin{align}
P(C_{lri} = 0 | c_r) &= 1-c_r\nonumber\\
P(C_{lri} = 1 | c_r) &= c_r
\end{align}
independent of the locus.

\subsection*{Genotype model}
We estimate the genotype of a locus given the conditional-SFS entry $Z_l=k$, $F_k$ is the probability that both alleles are IBD, and $\tau_k$ is the probability that the individual has a derived allele at position $k$
Thus
\begin{align}
P(G_l = 0| Z_l=k, \tau_k, F_k) &= F_k (1-\tau_k) + (1-F_k) (1-\tau_k)^2\nonumber\\
P(G_l = 1| Z_l=k, \tau_k, F_k) &= 2(1-F_k) \tau(1-\tau_k)\nonumber\\
P(G_l = 2| Z_l=k, \tau_k, F_k) &= F_k \tau_k + (1-F_k) \tau_k^2\label{eq:prg}
\end{align}

Estimating all the $\tau_k$ is one of the main goals of \texttt{admixslug}, as they can be used to calculate $F$-statistics and other quantities of interest.

For example, if we compare with the Altai Neandertal and Denisova 3 genomes, we would have the following $Z$ states:
\begin{table}[!ht]
	\begin{center}
		
		\begin{tabular}{r r r }
			$Z_l=k$ & Altai & Denisova 3 \\
			0 & 0 & 0 \\
			1 & 0 & 1 \\
			2 & 0 & 2 \\
			3 & 1 & 0 \\
			4 & 1 & 1 \\
			5 & 1 & 2 \\
			6 & 2 & 0 \\
			7 & 2 & 1 \\
			8 & 2 & 2 \\
		\end{tabular}
	\end{center}
\end{table}

This is implemented in \texttt{p\_gt\_diploid} and tested in \texttt{tests/test\_slug.py:test\_slug\_p\_gt\_diploid}.

For comparison with genotype-based methods, we also implement a \texttt{--gtmode} flag that allows the user to disable  the genotype likelihood and contamination models and directly supply the genotypes, in which case this is the only part of the model that is retained.



\subsection*{Likelihood and parameter estimation}

\subsubsection*{Complete Data Likelihood}
As in most graphical models, optimizing the exact likelihood is intractable because we would need to jointly optimize all parameters. 

We therefore use an EM-algorithm to estimate parameters. We proceed by calculating the complete-data likelihood (where we assume the latent variables) are known, and then derived the $Q$-function for estimating parameters. 

The complete data likelihood can be written as 
\begin{align*}
	\log P(\mathbf{O}, \mathbf{X}, \mathbf{C}, \mathbf{G} | \theta, \psi, \mathbf{Z}) &= 
	\sum_{lri} \log P(O_{lri} | X_{lri}, e, b)\\
	&+ \sum_{lri}\log P(C_{lri} | c_r)\\
	&+ \sum_{lri}\log P(X_{lri} | C_{lri}, \psi_l, G_l)\\
	&+ \sum_{l} \log P( G_l | Z_l, \tau_k, F_k)
\end{align*}

The corresponding Q-function is \begin{align*}
	Q(\theta | \theta') &= \mathbb{E}\big[\log P(\mathbf{O}, \mathbf{X}, \mathbf{C}, \mathbf{G} | \theta, \psi, \mathbf{Z}) | P(\mathbf{O, X, C, G}) | \theta', \mathbf{Z} \big]\\ &= 
	\sum_{lri} \log P(O_{lri} | X_{lri}, e, b) P(\mathbf{X} | \theta')\\
	&+ \sum_{lri}\log P(C_{lri} | c_r) P(\mathbf{C} | \theta')\\
	&+ \sum_{lri}\log P(X_{lri} | C_{lri}, \psi_l, G_l)P(\mathbf{C,G} | \theta')\\
	&+ \sum_{l} \log P( G_l | Z_l, \tau_k, F_k) P(\mathbf{G} | \theta')
\end{align*}

The probability terms are given in the previous section. What is missing is the expectation of the latent variables after each iteration. We calculate these recursively, using ``forward''-probabilities to calculate the probabilities of states given the parameters, and the ``backwards''-probabilities to calculate the probability of the observation given the latent states.



\subsection*{Forward Probabilities}
$P(G_l |Z_l )$ and $P(C_{rli} | c_r)$ can be calculated without knowing previous states.

\paragraph{Read probabilities}
For the read probabilities, we condition on $C$ and $G$:
\begin{align*}
	Pr(X_{lrj} | C_{lrj}=0) =& \sum_g Pr(X_{lrj} | C_{lrj}=0, G_l=g)Pr(G_l=g | Z_l) \\
    P(X_{lrj} | Z_l, \psi_l) =& Pr(X_{lrj} | C_{lrj}=0) Pr( C_{lrj}=0 | c_r) Pr(G_l|Z_l)\\ &+  P(X_{lrj} | C_{lrj}=1) Pr(C_{lrj}=1 | c_r) \\
\end{align*}

\subsection*{Backward Probabilities}
The elementary probabilities $Pr(\BFO|\BFX)$ can be calculated directly.

Given those, we can calculate the further probabilties for the other latent variables $\BFC$ and $\BFG$:

\begin{align*}
P(O_{l} | C_l) &= \prod_{rj} P(O_{lrj} | C_l)\\
&= \prod_{rj} P(O_{lrj} | X_{lrj}) P(X_{lrj} | C_l)\\
\end{align*}

\begin{align*}
	P(O_{lrj} | G_l) &=  P(O_{lrj} , C_{lrj}=0) + P(O_{lrj} , C_{lrj}=1, G_l)\\
	&= P(O_{lrj}| C_{rj}=0)Pr(C_{rj}=0) + \sum_x P(O_{lrj}| X_{lrj}=x) Pr(X_{lrj}=x | G_l, C_{rj}=1)\\\\
\end{align*}



\subsection*{Posteriors}
Using the forward and backward probabilities, we can calculate the posterior expectations for the latent variable in each E-step.


\paragraph{Posterior Genotypes}


The probability that genotype $G_l$ is 0, 1, 2 is 

\begin{align}
P(G_l=g|\BFO) &= \frac{P(\BFO|G_l=g)P(G_l=g|Z_l)}{P(G_l=g|Z_l)} \\	
&\propto  P(G_l | Z_l) \times \prod_{rj} P(O_{lrj} | G_l)
\end{align}

We ignore the denominator by calculating the probabilities for all three states and normalizing them to one, and all observation not associated with locus $l$ are the same for all genotypes, i.e. cancel out.


\paragraph{Posterior Reads}
The probability that read $X_{lrj}$ carries a derived allele
$$P(X_{lrj} | \BFO) \propto P(X_{lrj} | \BFC, \BFG) P(O_{lrj} | X_{lrj})$$

\paragraph{Posterior Contamination}
Calculate the posterior probability that read $rlj$ is contamination
$$P(C_{rlj}| \BFO)  \propto P(O_{rlj}|C_{rli}) c_r $$



\subsection{M-step}
Given the posterior expectations calculated in the E-step, we update the parameters for the next iteration in the $M$-step:
\subsubsection*{Estimating $e$ and $b$}
To simplify notation, let $n_{a,b,c}=\sum P(X_{lrj}=b|O_{lrj}=a, W_l=c)$.
\begin{align*}
\hat{e} & = \frac{n_{1,0,0} + n_{1,1,1}}{n_{1,0,0} + n_{1,1,1} + n_{0,0,0} +n_{0, 1, 1}}\\
\hat{b} & = \frac{n_{0,1,0} + n_{0, 0,1}}{n_{0, 1, 0} + n_{0, 0, 1} + n_{1, 1, 0} +n_{1, 0, 1}}
\end{align*}

For example, $n_{1,1,1} = Pr(X_{lrj}=1|O_{lrj}=1, W_l=1)$, i.e. a case where the read has a derived allele ($X_{lrj}=1)$, we observe an alternative allele ($O_{lrj}=1)$) and the locus is defined os that the alternative allele is the ancestral allele ($W_l=1$). This configuration happens if there is an error away from the reference, which is why this term appears in the estimation of the new $e$.

\subsubsection*{Estimating $c_k$}
\begin{align*}
c_r =  \sum_{li} P(C_{lri}| O_{lrj})
\end{align*},
i.e. we average over the posterior contamination estimates from all reads in the read group
\subsubsection*{Estimating $\tau_k$ and $F_k$}
Done numerically by optimizing
\begin{align*}
Q(\tau_k, F_k | \tau_k', F_k') 
&= \sum_{l} I[Z_l=k] \log P( G_l | Z_l, \tau_k, F_k) P(\mathbf{G} | \tau_k', F_k') \\
&= \sum_{l} I[Z_l=k]\sum_{g=0}^2 \log P( G_l=g | Z_l=k, \tau_k, F_k)P(G=g | \tau_k', F_k')
\end{align*}
where $I$ is an indicator function, $P(G=g | \tau_k', F_k')$ are the estimates from the previous iteration and $ P( G_l=g | Z_l=k, \tau_k, F_k)$ are given by eq \ref{eq:prg}

\subsection*{Calculating F-statistics}
We can calculate \emph{some} $F$-statistics directly from the \texttt{admixslug} output. We use the estimates based on pairwise differences \citep{peter2016}:
\begin{align}
F_2(X, Y) &= 2\pi_{xy} -\pi_{xx} - \pi_{yy}\\
F_3(X; Y, Z) &= \pi_{xy} +\pi_{xz} - \pi_{yz} - \pi_{xx}\\
F_4(X, Y; Z, W) &= \pi_{xz} +\pi_{yw} - \pi_{xw} - \pi_{yz}.
\end{align}
Assume we have $L$ loci, and population $a_{il}, d_{il}$ are the ancestral/derived counts in population $i$ at locus $l$, respectively, such that $n_{il} = a_{il} + d_{il}$. Then
\begin{equation}
\pi_{ij} = \begin{cases}
\frac{1}{L}\sum_l \frac{a_{il}d_{jl} + d_{il}a_{jl}} {n_{il}n_{jl}} ,& \text{if } i \neq j\\
\frac{1}{L}\sum_l \frac{a_{il}d_{il}}{n_{il}(n_{il} -1)},              & \text{if }i =j
\end{cases}
\end{equation}

using the conditonal SFS from \texttt{admixslug}, we can write equivalently
\begin{equation}
\pi_{ij} = \begin{cases}
\sum_k s_k \frac{a_{ik}d_{jk} + d_{ik}a_{jk}} {n_{ik}n_{jk}} ,& \text{if } i \neq j\\
\sum_k s_k \frac{a_{ik}d_{ik}}{n_{ik}(n_{ik} -1)},              & \text{if }i =j
\end{cases},
\end{equation}
where $a_{ik}, d_{ik}, n_{ik}$ are now the counts of ancestral/derived/total alleles in population $i$ and SFS-category $k$, and $s_k$ is the (estimated) proportion of SNPs of this category.

These equations can directly be used to calculate $\pi$ within and between pairs of reference populations. To calculate $\pi$ between a reference population and the target individual, we use the estimator

\begin{equation}
\pi_{is} = \sum_k \frac{s_k}{n_{ik}} \left[\tau_k a_{ik}+ (1-\tau_k) d_{ik}\right],
\end{equation}
since $\tau_k$ is the expected proportion of SNPs carrying a derived allele in SFS-category $k$.


One caveat is that we cannot calculate $\pi{ss}$, the heterozygosity in the target individual. Also, for references without heterozygosity (e.g. the chimp-outgroup, or pseudo-haploid reference individuals), $\pi_{ii}$ cannot be calculated. By convention, we set these to zero. Thus, $F$-statistics involving these heterozygosities ($F_2$-statistics, and $F_3$-statistics using these individuals as samples) will be overestimated by a constant.


\section{Calculating F(A|B) statistics}
The $F(A|B)$ statistics where $B$ is a single diploid genome are the proportion of derived alleles at sites where $B$ is heterozygous. This corresponds exactly to the $\tau_k$ for categories where $k$ is heterozygous.

Thus,

\begin{equation}
	F(A|B) = \frac{\sum_k s_k \tau_k}{\sum_k s_k},
\end{equation}

where the sum is only over entries where $B$ is heterozygous.

\bibliography{main}
\bibliographystyle{plainnat}


\end{document}
